---
title: "mishmashr: A Unified Interface for Fitting Mash and Flash Models"
author: 
  - Eric Weine^[University of Chicago, ericweine15@gmail.com]
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{mishmashr: A Unified Interface for Fitting Mash and Flash Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Mathematical Framework

The creation of this package was inspired by the mathematical similarities between flash and mash models. Suppose $\hat{\beta}$ is an $n \times p$ matrix of observations. We assume the rows of $\hat{\beta}$ are independent, and

\begin{equation*}
    \hat{\beta}_{i\cdot} \sim N_{p}\big(\beta_{i\cdot}, V^{(i)}\big)
\end{equation*}
where $\hat{\beta}_{i\cdot}$ denotes the $i^{th}$ row of $\hat{\beta}$, $N_{p}$ denotes the $p$ dimensional multivariate normal distribution and $V^{(i)}$ is a known $p \times p$ covariance matrix. Now, if we would like to impose a "flash" prior on $\beta$, we could assume that each row of $\beta$ is generated independently as 
\begin{equation*}
    \beta_{i\cdot} \sim N_{p}\big(\boldsymbol{\ell}_{i}F^{T}, diag(\sigma^{2}_{i1}, \dots, \sigma^{2}_{ip})\big)
\end{equation*}
where each $\boldsymbol{\ell}_{i}$ is an n-vector of unknown "loadings", $F$ is an $p \times k$ matrix of unknown "factors", and $\sigma^{2}_{i1}, \dots, \sigma^{2}_{ip}$ are unknown variance parameters. Alternatively, if we would like to impoase a "mash" prior on $\beta$, we could assume that each row of $\beta$ is generated independently as 
\begin{equation*}
    \beta_{i\cdot} \sim \sum_{k = 1}^{K} \sum_{l = 1}^{L} N_{p}\big(\boldsymbol{0}, \omega_{l}U_{k}\big)
\end{equation*}

where each $\omega_{l}$ is a scaling factor and each $U_{k}$ is an unknown covariance matrix. Thus, the conceptual difference between mash and flash can be viewed (in part) as a difference in choice of prior for each row of $\beta$. Note that while mash can handle any (positive semi-definite) choice for each $V^{(i)}$, flash is currently constrained to only handle diagonal choices for each $V^{(i)}$. Thus, ``mishmashr`` only handles diagonal observation covariance matrices.

## Flash Enhancements

While the formulation of flash above is useful for comparing it to mash, for the remainder of this vignette we will return to the more familiar formulation of

\begin{equation*}
Y = LF^{T} + E^{(1)} + E^{(2)}
\end{equation*}

where $L$ is an $n \times k$ matrix of loadings, $F$ is a $p \times k$ matrix of factors, $E^{(1)}$ is an $n \times p$ matrix where each $e^{(1)}_{ij}$ is generated independently as $N(0, \sigma_{ij}^{2})$ where $\sigma_{ij}$ is unknown, and $E^{(2)}$ is an $n \times p$ matrix where each $e^{(2)}_{ij}$ is generated independently as $N(0, s_{ij}^{2})$ where $s_{ij}$ is known.

### lfsr

Currently, `flashier` implements lfsr calculations for $L$ and $F$, but contains no lfsr calculation for $LF^{T} + E^{(1)}$, which could be of interest. To obtain such values, `mishmashr` uses a two-step simulation process.

(1) Sample $M$ times from $f(LF^{T}|Y)$.

(2) For each sample $m$, sample once from $f(E^{(1)}|(LF^{T})^{(m)}, \sigma^{2}_{ij})$ and add it to $(LF^{T})^{(m)}$. 

Step 1 is already implemented in flashier, but step 2 is not. To implement step 2, we consider the distribution of residual $(Y - (LF^{T})^{(m)})_{ij} = r_{ij}$ given $L$, $F$, and $\sigma^{2}_{ij}$. Then, we can write 
\begin{align*}
r_{ij} &\sim N(e^{(1)}_{ij}, s_{ij}^{2})\\
e_{ij} & \sim N(0, \sigma_{ij}^{2}).
\end{align*}

It is a well known fact that 

\begin{equation*}
f(e^{(1)}_{ij} | r_{ij}) \sim N(\mu_{ij}, \gamma_{ij})
\end{equation*}
where $\mu_{ij} = \frac{\frac{1}{s_{ij}^{2}}r_{ij}}{\frac{1}{s_{ij}^{2}} + \frac{1}{\sigma_{ij}^{2}}}$ and $\gamma_{ij} = \frac{1}{\frac{1}{s_{ij}^{2}} + \frac{1}{\sigma_{ij}^{2}}}$. In practice, $\sigma_{ij}^{2}$ is not known and instead will be replaced with its MLE from fitting flash.

### Expectation of E

To derive $E[e^{(1)}_{ij}]$, assuming $\sigma^{2}_{ij}$ known, we simply note that

\begin{align*}
E[e_{ij}^{(1)}] &= E[E[e_{ij}^{(1)} | \boldsymbol{\ell}_{i}, \boldsymbol{f}_{j}]] \\
&= E\Bigg[\frac{\frac{1}{s_{ij}^{2}}r_{ij}}{\frac{1}{s_{ij}^{2}} + \frac{1}{\sigma_{ij}^{2}}}\Bigg] \\
&= \frac{\frac{1}{s_{ij}^{2}}E[r_{ij}]}{\frac{1}{s_{ij}^{2}} + \frac{1}{\sigma_{ij}^{2}}} \\
&= \frac{\frac{1}{s_{ij}^{2}}(y_{ij} - E[\boldsymbol{\ell}_{i} \cdot \boldsymbol{f}_{j}])}{\frac{1}{s_{ij}^{2}} + \frac{1}{\sigma_{ij}^{2}}} \\
\end{align*}
which is easily computed, since `flashier` calculates $E[\boldsymbol{\ell}_{i} \cdot \boldsymbol{f}_{j}]$.

### Mashy Flash

The two enhancements above are implmented in `mishmashr::mashy_flash`, named because it gives a mash-like interface for fitting a flash model.

First, we'll generate some data from a low-rank covariance structure.

```{r, eval=F}
# Generate 10 x 500 matrix with rank 3.
n <- 10
p <- 100
k <- 3

A <- matrix(
  data = rnorm(n = n * k, sd = 2), nrow = n, ncol = k
)
cov <- A %*% t(A) / (n - 1)

X <- t(MASS::mvrnorm(n = p, mu = rep(0, n), Sigma = cov))

# N(0, 1) noise for each element of E1
E1 <- matrix(rnorm(n * p, 0, 1), n, p)

# N(0, 1) noise for each element of E2
E2 <- matrix(rnorm(n * p, 0, 1), n, p)

Y <- X + E1 + E2
```

Now, we can create a mash data object.

```{r, eval=F}
mash_data <- mashr::mash_set_data(Bhat = Y, Shat = 1)
```

Finally, we can pash this data into `mashy_flash`, which estimates a flash model using the mash data. Any other parameters which can be passed to `flashier::flash` can be passed to `mashy_flash`.

```{r, eval=F}
fl <- mishmashr::mashy_flash(
  mash_data,
  greedy.Kmax = 5, 
  backfit = TRUE,
  verbose = 0
)
```

`mashy_flash` outputs a `flashier::flash` object, but with a few extra goodies. It contains an element `LF_E.lfsr`, which contains the lfsr of `LF^{T} + E`. It also contains an element `E.pm`, containing $E[E^{(1)} | Y]$. We've also implemented the function `get_flash_pm`, which returns $E[LF^{T} + E^{(1)} | Y]$ from a flash object.

## Unified Interface

In addition to the enhancements to flash models, `mishmashr` implements a unified interface, `mishmashr::mishmash` for fitting flash and mash models. `mishmash` simply takes a mash data object and two optional lists of parameters to `mashr::mash` and `flashier::flash`.

```{r, eval=FALSE}
Ulist = mashr::cov_canonical(mash_data)

mv_models <- mishmashr::mishmash(
  mash_data = mash_data,
  mash_params = list(Ulist = Ulist, posterior_samples = 100),
  flash_params = list(greedy.Kmax = 5, backfit = TRUE, verbose = 0)
)
```

After fitting both mash and flash models, `mv_models` will have two elements,
`flash` and `mash`, which contain each of the fit objects.

## Session Info

```{r session-info}
sessionInfo()
```

